---
 title: "Finding Coexpression Modules"
author: "Gokul Seshadri"
date: "2025-04-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Importing the necessary libraries

```{r}
library(tibble)
library(dplyr)
library(ggplot2)
library(pheatmap)
library(WGCNA)
library(glue)
library(reshape2)
library(enrichR)
library(clusterProfiler)
library(org.Hs.eg.db)
library(ReactomePA)

enableWGCNAThreads(nThreads = 8)
```

### Loading the dataset having CMV significant genes

```{r}
log2cpm_df = read.csv("D:/RA/GitHub/Plan-B/Datasets/Final Datasets/CMV_sig_data_log2cpm.csv")

# selecting only the CMV significant genes
only_genes_df = log2cpm_df[, grep("^ENSG", names(log2cpm_df))]
rownames(only_genes_df) = log2cpm_df$Lab_ID
rm(log2cpm_df)

dim(only_genes_df)
```

<br>

### Sample clustering and removing outlier samples
```{r}
#Group data in a dendogram to check outliers
sampleTree = hclust(dist(only_genes_df), method = "average")
```

```{r}
par(cex = 0.6)
plot(sampleTree, main = "Sample clustering to detect outliers", sub="", xlab="", cex.lab = 1.5, 
     cex.axis = 1.5, cex.main = 2)

#Plot a line showing the cut-off
abline(h = 100, col = "red")
```

```{r}
#Determine clusters below the line
clust = cutreeStatic(sampleTree, cutHeight = 100, minSize = 10)

#Cluster 1 contains the samples we want to keep.
keepSamples = (clust==1)
only_genes_df2 = only_genes_df[keepSamples, ]

nGenes = ncol(only_genes_df2)
nSamples = nrow(only_genes_df2)

# after filteration based on Tree height
# only_genes_df = only_genes_df[-31, ] # F1652275 index sample was outlier
dim(only_genes_df2)
```

<br>

#### Filtering out low-variance genes before running WGCNA
```{r}
keepGenes <- apply(only_genes_df2, 2, var) > quantile(apply(only_genes_df2, 2, var), 0.25)
only_genes_filtered <- only_genes_df2[, keepGenes]

dim(only_genes_filtered)
```


### Adding the clinical data
```{r}
clinical_df = read.csv("D:/RA/GitHub/Plan-B/Datasets/Final Datasets/CMV_clonotypes_lv.csv")
clinical_df = clinical_df[, c('Lab_ID', 'CD4N', 'TRAB', 'age')] %>%
                column_to_rownames("Lab_ID")

clinical_df = clinical_df[keepSamples, ]
dim(clinical_df)
```

<br>

### Network construction
```{r}
# allowWGCNAThreads()

orig_cor <- cor       
cor <- WGCNA::cor

# Choose a set of soft-thresholding powers
powers = c(c(1:10), seq(from = 12, to = 20, by = 2))

# Call the network topology analysis function
sft = pickSoftThreshold(
  only_genes_filtered,             
  #blockSize = 30,
  corFnc = cor,
  powerVector = powers,
  verbose = 5
  )

par(mfrow = c(1,2));
cex1 = 0.9;

# Scale-free topology fit index as a function of the soft-thresholding power
plot(sft$fitIndices[, 1],
     -sign(sft$fitIndices[, 3]) * sft$fitIndices[, 2],
     xlab = "Soft Threshold (power)",
     ylab = "Scale Free Topology Model Fit, signed R^2",
     main = paste("Scale independence")
)
text(sft$fitIndices[, 1],
     -sign(sft$fitIndices[, 3]) * sft$fitIndices[, 2],
     labels = powers, cex = cex1, col = "red"
)
abline(h = 0.90, col = "red")

# Mean connectivity as a function of the soft-thresholding power
plot(sft$fitIndices[, 1],
     sft$fitIndices[, 5],
     xlab = "Soft Threshold (power)",
     ylab = "Mean Connectivity",
     type = "n",
     main = paste("Mean connectivity")
)
text(sft$fitIndices[, 1],
     sft$fitIndices[, 5],
     labels = powers,
     cex = cex1, col = "red")
```


```{r}
enableWGCNAThreads(nThreads = 12)

softPower = 8;
adjacency = adjacency(only_genes_filtered, power = softPower)

# Turn adjacency into topological overlap
TOM = TOMsimilarity(adjacency)

# Finding TOM dissimilarity
dissTOM = 1-TOM
```

#### Hclust with TOM based dissimilarity
```{r}
geneTree = hclust(as.dist(dissTOM), method = "average")

# Plot the resulting clustering tree (dendrogram)
plot(geneTree, xlab="", sub="", main = "Gene clustering on TOM-based dissimilarity",
     labels = FALSE, hang = 0.04)
```

#### Partition genes with similar expression into modules
```{r}
# We like large modules, so we set the minimum module size relatively high:
minModuleSize = 100;
# Module identification using dynamic tree cut:
dynamicMods = cutreeDynamic(dendro = geneTree, distM = dissTOM,
                deepSplit = 2, pamRespectsDendro = FALSE,
                minClusterSize = minModuleSize);

table(dynamicMods)
```

```{r}
dynamicColors = labels2colors(dynamicMods)

# Plot the dendrogram and colors underneath
plotDendroAndColors(geneTree, dynamicColors, "Dynamic Tree Cut",
                    dendroLabels = FALSE, hang = 0.03,
                    addGuide = TRUE, guideHang = 0.05,
                    main = "Gene dendrogram and module colors")
```


#### Cluster and merge modules based on eigen genes

```{r}
# Calculate eigengenes
MEList = moduleEigengenes(only_genes_filtered, colors = dynamicColors)
MEs = MEList$eigengenes
# Calculate dissimilarity of module eigengenes
MEDiss = 1-cor(MEs);
# Cluster module eigengenes
METree = hclust(as.dist(MEDiss), method = "average");
```

```{r}
# Plot the result
plot(METree, main = "Clustering of module eigengenes",
     xlab = "", sub = "")
MEDissThres = 0.2
# Plot the cut line into the dendrogram
abline(h=MEDissThres, col = "red")
```

```{r}
# Call an automatic merging function
merge = mergeCloseModules(only_genes_filtered, dynamicColors, cutHeight = MEDissThres, verbose = 3)
```

```{r}
# The merged module colors
mergedColors = merge$colors;
# Eigengenes of the new merged modules:
mergedMEs = merge$newMEs;

#pdf(file = "Plots/geneDendro-3.pdf", wi = 9, he = 6)
plotDendroAndColors(geneTree, cbind(dynamicColors, mergedColors),
                    c("Dynamic Tree Cut", "Merged dynamic"),
                    dendroLabels = FALSE, hang = 0.03,
                    addGuide = TRUE, guideHang = 0.05)
```

```{r}
table(mergedColors)
```

```{r}
# Rename to moduleColors
moduleColors = mergedColors

# Construct numerical labels corresponding to the colors
colorOrder = c("grey", standardColors(50));
moduleLabels = match(moduleColors, colorOrder)-1;
MEs = mergedMEs;

# Save module colors and labels for use in subsequent parts
# save(MEs, moduleLabels, moduleColors, geneTree, file = "D:/MS/My Research/Plan B/Code/Coexpression Analysis/cpm3/cpm3_netCon_stepByStep.RData")
```

#### saving modular info in a csv file
```{r}
# creating a dataframe to store module information
coexp_modules = data.frame(gene_id = names(only_genes_filtered), mod_num = moduleLabels, mod_color = moduleColors)
coexp_modules$gene_id = sub("\\..*$", "", coexp_modules$gene_id)
coexp_modules
```

```{r}
write.csv(coexp_modules, 'D:/RA/GitHub/Plan-B/Code/Coexpression Analysis/coexp_modules.csv', row.names = FALSE)
```

### Study Relationships between modules and traits
```{r}
# Define numbers of genes and samples
nGenes = ncol(only_genes_filtered);
nSamples = nrow(only_genes_filtered);

# Recalculate MEs with color labels
MEs0 = moduleEigengenes(only_genes_filtered, moduleColors)$eigengenes
MEs = orderMEs(MEs0)
moduleTraitCor = cor(MEs, clinical_df, use = "p");
moduleTraitPvalue = corPvalueStudent(moduleTraitCor, nSamples);


# Will display correlations. For space reasons, we will not display the p values of the correlations. 
textMatrix =  signif(moduleTraitCor, 2)
textMatrix = formatC(moduleTraitCor, digits = 2, format = "f")

dim(textMatrix) = dim(moduleTraitCor)

# Display the correlation values within a heatmap plot
labeledHeatmap(Matrix = moduleTraitCor,
               xLabels = names(clinical_df),
               yLabels = names(MEs),
               ySymbols = NULL,
               colorLabels = FALSE,
               colors = blueWhiteRed(50),
               textMatrix = textMatrix,
               setStdMargins = FALSE,
               cex.text = 1,
               zlim = c(-1,1), 
               main = paste("Gene Expression Module- Trait relationships"), ylab = "Gene Expression-Based Modules")
```


### Saving the results in Cytoscape format
```{r}
# saving info about the undirected edges
nTopEdges = floor(length(TOM) * 0.05)

edges = data.frame(
  from = rep(coexp_modules$gene_id, each = length(coexp_modules$gene_id)),
  to = rep(coexp_modules$gene_id, times = length(coexp_modules$gene_id)),
  weight = as.vector(TOM)
)
edges = edges[edges$from != edges$to, ]
edges = edges[order(-edges$weight), ][1:nTopEdges, ]

write.table(edges, file = "D:/RA/GitHub/Plan-B/Code/Coexpression Analysis/all_modules_edges.txt", sep = "\t", row.names = FALSE, quote = FALSE)

# saving info about the nodes
nodeAttr = data.frame(
  id = coexp_modules$gene_id,
  moduleColor = coexp_modules$mod_color,  # from WGCNA output
  stringsAsFactors = FALSE
)
write.table(nodeAttr, file = "D:/RA/GitHub/Plan-B/Code/Coexpression Analysis/all_modules_nodes.txt", sep = "\t", row.names = FALSE, quote = FALSE)
```

### Creating dataset for downstream analysis
```{r}
### Gene Expression data
log2cpm_df = read.csv("D:/MS/My Research/Plan B/Datasets/Final Datasets/CMV_sig_data_log2cpm.csv")
names(log2cpm_df) = sub("\\..*$", "", names(log2cpm_df))

# selecting only gene related columns
log2cpm_df = log2cpm_df[, c("Lab_ID", coexp_modules$gene_id)]

### Data with latent variable info
clinical_lv = read.csv("D:/MS/My Research/Plan B/Datasets/Final Datasets/CMV_clonotypes_lv.csv")

### Combining the two
latent_mod_data = inner_join(clinical_lv, log2cpm_df, by = "Lab_ID")
latent_mod_data = latent_mod_data[keepSamples, ]
dim(latent_mod_data)
```

```{r}
write.csv(latent_mod_data, "D:/RA/GitHub/Plan-B/Datasets/Final Datasets/final_dataset.csv", row.names = FALSE)
```


### My previous Method
Forming modules,
```{r}
enableWGCNAThreads(nThreads = 12)

picked_power = 8

netwk <- blockwiseModules(only_genes_filtered,                

                          # == Adjacency Function ==
                          power = picked_power,
                          TOMType = "unsigned",
                          networkType = "unsigned",

                          # == Tree and Block Options ==
                          deepSplit = 2,
                          pamRespectsDendro = T,
                          minModuleSize = 30,
                          maxBlockSize = 10000,
                          
                          # == Module Adjustments ==
                          reassignThreshold = 0,
                          mergeCutHeight = 0.25,
                          
                          # == TOM == Archive the run results in TOM file (saves time)
                          saveTOMs = T,
                          saveTOMFileBase = "SpodopteraTOM-blockwise",
                          
                          # == Output Options
                          numericLabels = T,
                          verbose = 3)
```

```{r}
# Convert labels to colors for plotting
mergedColors = labels2colors(netwk$colors)
names(mergedColors) = names(netwk$colors)
  
# Plot the dendrogram and the module colors underneath
plotDendroAndColors(
  netwk$dendrograms[[1]],
  mergedColors[netwk$blockGenes[[1]]],
  "Module colors",
  dendroLabels = FALSE,
  hang = 0.03,
  addGuide = TRUE,
  guideHang = 0.05 )
```

```{r}
# creating a dataframe to store module information
coexp_modules = as.data.frame(netwk$colors) %>%
                  rownames_to_column("gene_id") %>%
                  dplyr::rename(modules = "netwk$colors")
#coexp_modules$gene_id = sub("\\..", "", coexp_modules$gene_id)
coexp_modules$colors = labels2colors(coexp_modules$modules)
coexp_modules
```

```{r}
write.csv(coexp_modules, 'D:/MS/My Research/Plan B/Code/Coexpression Analysis/coexp_modules.csv', row.names = FALSE)
```

### Pathway Enrichment Analysis for WGCNA modules

```{r}
# Creating a list of modules
modules = unique(coexp_modules$mod_color)  # Get unique module colors

moduleGenes = lapply(modules, function(module) {
   coexp_modules$gene_id[coexp_modules$mod_color == module] # Get genes in each module
})

names(moduleGenes) = modules  # Assign module colors as names


# modules = unique(mergedColors)  # Get unique module colors
# 
# moduleGenes = lapply(modules, function(module) {
#   sub("\\..", "", names(mergedColors)[mergedColors == module])  # Get genes in each module
# })
# 
# names(moduleGenes) = modules  # Assign module colors as names
```


```{r}
# Converting ENSEMBL ids to ENTREZIDs
convertToEntrez <- function(genes) {
  mappedIDs = mapIds(org.Hs.eg.db, keys=genes, column="ENTREZID", keytype="ENSEMBL", multiVals="first")
  return(na.omit(mappedIDs))  # Remove missing IDs
}

moduleGenesEntrez = lapply(moduleGenes, convertToEntrez)
```

```{r}
# GO Enrichment Analysis
goResults = lapply(moduleGenesEntrez, function(genes) {
  enrichGO(gene          = genes,
           OrgDb         = org.Hs.eg.db,
           keyType       = "ENTREZID",
           ont           = "BP",  # Biological Process
           pAdjustMethod = "fdr",
           pvalueCutoff  = 0.05)
})

# KEGG Pathway Enrichment
keggResults = lapply(moduleGenesEntrez, function(genes) {
  enrichKEGG(gene          = genes,
             organism      = "hsa",  # Human
             keyType       = "kegg",
             pAdjustMethod = "fdr",
             pvalueCutoff  = 0.05)
})

# Reactome Pathway Enrichment
reactomeResults = lapply(moduleGenesEntrez, function(genes) {
  enrichPathway(gene          = genes,
                organism      = "human",
                pAdjustMethod = "fdr",
                pvalueCutoff  = 0.05)
})
```


```{r}
# helper function to print pathways ordered by gene ratio
print_pathways <- function(Results, module="", pathway="Go", show_n=10){
  # Extract the enrichment results as a data frame
  go_df <- as.data.frame(Results)
  
  # Convert GeneRatio (which is in "x/y" format) to numeric
  go_df$GeneRatioValue <- sapply(go_df$GeneRatio, function(x) {
    eval(parse(text = x))
  })
  
  # Take top 25 GO terms by GeneRatio
  top_go <- go_df %>%
    arrange(desc(GeneRatioValue)) %>%
    slice_head(n = show_n)
  
  # Make Description a factor ordered by GeneRatioValue for plotting
  top_go$Description <- str_wrap(top_go$Description, width = 40)
  top_go$Description <- fct_reorder(top_go$Description, top_go$GeneRatioValue)
  top_go$GeneRatioPercent <- top_go$GeneRatioValue * 100
  top_go$GeneRatioPercent <- round(top_go$GeneRatioPercent, 2)
  
  # Plot using ggplot2
  ggplot(top_go, aes(x = GeneRatioPercent, y = Description)) +
    geom_point(aes(size = Count, color = -log10(p.adjust))) +
    scale_color_gradient(low = "blue", high = "red") +
    #scale_x_continuous(breaks = pretty(top_go$GeneRatioPercent, n = 5)) +
    labs(
      title = glue("{pathway} terms for {module}"),
      x = "Gene Ratio (%)",
      y = "",
      color = "-log10(FDR)",
      size = "Gene Count"
    ) +
    theme_minimal(base_size = 14) +
    theme(axis.text.y = element_text(size = 12))
}

```


```{r}
#dotplot(goResults$black, showCategory=10, title="GO Enrichment for Black Module")
print_pathways(goResults$black, module="Black", pathway="GO")
print_pathways(keggResults$black, module="Black", pathway="KEGG")
print_pathways(reactomeResults$black, module="Black", pathway="Reactome")
```

```{r}
print_pathways(goResults$blue, module="Blue", pathway="GO")
print_pathways(keggResults$blue, module="Blue", pathway="KEGG")
print_pathways(reactomeResults$blue, module="Blue", pathway="Reactome")
```

```{r}
print_pathways(goResults$brown, module="Brown", pathway="GO")
#dotplot(keggResults$brown, showCategory=10, title="KEGG Enrichment for Brown Module")
#dotplot(reactomeResults$brown, showCategory=10, title="Reactome Enrichment for Brown Module")
```

```{r}
dotplot(goResults$magenta, showCategory=10, title="GO Enrichment for Magenta Module")
dotplot(keggResults$magenta, showCategory=10, title="KEGG Enrichment for Magenta Module")
dotplot(reactomeResults$magenta, showCategory=10, title="Reactome Enrichment for Magenta Module")
```

```{r}
print_pathways(goResults$pink, module="Pink", pathway="GO")
print_pathways(keggResults$pink, module="Pink", pathway="KEGG")
#dotplot(keggResults$pink, showCategory=10, title="KEGG Enrichment for Pink Module")
print_pathways(reactomeResults$pink, module="Pink", pathway="Reactome")
```

```{r}
dotplot(goResults$purple, showCategory=10, title="GO Enrichment for Purple Module")
dotplot(keggResults$purple, showCategory=10, title="KEGG Enrichment for Purple Module")
dotplot(reactomeResults$purple, showCategory=10, title="Reactome Enrichment for Purple Module")
```

```{r}
#dotplot(goResults$red, showCategory=10, title="GO Enrichment for Red Module")
#dotplot(keggResults$red, showCategory=10, title="KEGG Enrichment for Red Module")
#dotplot(reactomeResults$red, showCategory=10, title="Reactome Enrichment for Red Module")
```

```{r}
dotplot(goResults$turquoise, showCategory=10, title="GO Enrichment for Turquoise Module")
dotplot(keggResults$turquoise, showCategory=10, title="KEGG Enrichment for Turquoise Module")
dotplot(reactomeResults$turquoise, showCategory=10, title="Reactome Enrichment for Turquoise Module")
```

```{r}
#dotplot(goResults$yellow, showCategory=10, title="GO Enrichment for Yellow Module")
#dotplot(keggResults$yellow, showCategory=10, title="KEGG Enrichment for Yellow Module")
dotplot(reactomeResults$yellow, showCategory=10, title="Reactome Enrichment for Yellow Module")
```

```{r}
#dotplot(goResults$grey, showCategory=10, title="GO Enrichment for Grey Module")
dotplot(keggResults$grey, showCategory=10, title="KEGG Enrichment for Grey Module")
dotplot(reactomeResults$grey, showCategory=10, title="Reactome Enrichment for Grey Module")
```




