---
title: "Effect of CMV on TRAB clones"
author: "Gokul Seshadri"
date: "2025-02-13"
output:
  html_document:
    code_folding: hide
    smart: false
    theme: cerulean
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: false
    number_sections: true
    df_print: paged
---

```{r libraries, include=FALSE, echo=FALSE}
# invoking necessary packages
library("tidyverse")
library("tibble")
library("reshape2")
library("multcompView")
library("car")

library('broom.mixed')
library("ggplot2")
library("glue")
library("knitr")

library("tableone")
library("labelled")
library("MatchIt")
library("survey")

knitr::opts_chunk$set(echo = TRUE, message = FALSE ,warning= FALSE,
                      comment = "",fig.width = 12,fig.height = 8)
```


<div style="text-align: justify">
### **Data Preparation**
</div>

Reading the dataset and doing appropriate type conversions

```{r}
# reading the combined dataframe
data_df = read.csv('D:/RA/GitHub/Plan-B/Datasets/Final Datasets/CMV_clonotypes_lv.csv')

# converting blocks and treatment into factors
data_df$sex = factor(data_df$sex)
data_df$race = factor(data_df$race, levels=c("Non-Hispanic White", "Non-Hispanic Black", "Hispanic", "Non-Hispanic Other"))
data_df$smoke_stat = factor(data_df$smoke_stat, levels=c("Never smokers", "Former smokers", "Current smokers"))

# removing those with empty CMV status
data_df = data_df[data_df$CMV_status != '', ]
data_df$CMV_status = ifelse(data_df$CMV_status == "Positive", 1, 0)

# immune cell variables
clonotypes = c('TRA_total', 'TRB_total', 'TRG_total',
               'TRA_unique', 'TRB_unique', 'TRG_unique')
```

#### **Removing nulls**

```{r}
# null counts in each column
sapply(data_df, function(x) sum(is.na(x)))

# removing null rows in covariates
columns_to_check =c('CMV_status', 'age', 'sex', 'race', 'mortality', 'lymp', 'TRA_total', 'TRB_total', 'TRAB')
filtered_df <- data_df %>% 
                  filter(across(all_of(columns_to_check), ~ !is.na(.)))

cat("\nFiltered data size:", nrow(filtered_df))
```

```{r, echo=FALSE}
# setting variable names for TableOne
filtered_df = labelled::set_variable_labels(filtered_df,
                                      age = "Age at 2016",
                                      sex = "Biological Sex",
                                      race = "Racial Group",
                                      educ = "Education Years",
                                      smoke_stat = "Smoking Status",
                                      mortality = "Mortality",
                                      TRAB = "T alpha-beta clones",
                                      Tcells = "T cells count",
                                      CD4N_pct = "Naive Helper T cells pct",
                                      BMI = "Body Mass Index",
                                      INFLAM_5 = "Total Inflammation Score",
                                      comorb_idx = "# of comorbidities"
                                      )

# creating a table one stratified by treatment group
cat_cols = c('sex', 'race', 'mortality', 'smoke_stat')
num_cols = c('age', 'educ', 'BMI', 'INFLAM_5', 'comorb_idx', 'Tcells', 'TRAB', 'CD4N_pct')
all_cols = c(cat_cols, num_cols)

HRS_table = CreateTableOne(data = filtered_df, vars=all_cols, factorVars = cat_cols, strata = "CMV_status", test=TRUE)
HRS_table = print(HRS_table, nonnormal=num_cols, varLabels = TRUE);
```

```{r}
kable(HRS_table, caption = "**HRS Participant Characteristics**", align = "c")
```

<br>
<br>


### **Investigating CMV's impact on TRAB Clone**
#### **Finding the unadjusted mean difference**

```{r}
## finding un-adjusted ATE
TRAB_t = filtered_df$TRAB[filtered_df$CMV_status == 1]
TRAB_c = filtered_df$TRAB[filtered_df$CMV_status == 0]

ATE = mean(TRAB_t, na.rm = T) - mean(TRAB_c, na.rm = T)

# finding the weighted standard error
SE = sqrt(var(TRAB_t, na.rm = T)/length(TRAB_t) + var(TRAB_c, na.rm = T)/length(TRAB_c))

# finding the confidence interval
CI = ATE + c(-1, 1) * qnorm(0.975) * SE

# test for significance
p_val = t.test(TRAB_t, TRAB_c, var.equal = TRUE)$p.value
 
output1 <- data.frame(
  Trt_Effect = round(ATE, 4),
  Std_Err = round(SE, 4),
  lower_CI = round(CI[1], 4),
  upper_CI = round(CI[2], 4),
  p_value = round(p_val, 4)
)

kable(output1, caption = "Unadjusted ATE of CMV on TRAB", align = "c")
```

We can conclude that there is a statistically significant differences in the TRAB clone count between CMV positive and negative individuals.

<br>
<br>


#### **An adjusted model to find ATE and ATT**

Average Treatment Effect of CMV on TRAB clones
```{r}
# using the function to calculate ATE and CI from the homework 2 key
calculate_ATE_lm <- function(model_formula, data, num_bootstraps = 100) {
 
  model = lm(model_formula, data = data)
  
  data_trt = data_con <- data
  data_trt$CMV_status = 1
  data_con$CMV_status = 0
  
  pred1 = predict(model, newdata = data_trt, type = "response")
  pred0 = predict(model, newdata = data_con, type = "response")
  ATE = mean(pred1 - pred0, na.rm = TRUE)
  
  # Bootstrap replication for SE
  ate_boot <- c()
  set.seed(101)
  for (i in 1:num_bootstraps) {
    data_boot = data[sample(nrow(data), size = nrow(data), replace = TRUE), ]
    model_boot = lm(model_formula, data = data_boot)
    
    data_trt_boot = data_con_boot <- data_boot
    data_trt_boot$CMV_status = 1
    data_con_boot$CMV_status = 0
    
    pred1_boot = predict(model_boot, newdata = data_trt_boot, type = "response")
    pred0_boot = predict(model_boot, newdata = data_con_boot, type = "response")
    ate_boot = c(ate_boot, mean(pred1_boot - pred0_boot, na.rm = TRUE))
  }
  
  # Calculate standard error and confidence interval
  SE = sd(ate_boot, na.rm = TRUE)
  CI = ATE + c(-1, 1) * qnorm(0.975) * SE
  CI_str = paste0(round(CI[1], 4), ", ", round(CI[2], 4))
  
  return(list(ATE = round(ATE, 4), SE = round(SE, 4), CI = CI_str))
}


# the average treatment effect (ATE)
model_formula = TRAB ~ CMV_status * age + sex + race + educ + smoke_stat + INFLAM_5 + comorb_idx + Tcells
result_1 = calculate_ATE_lm(model_formula, filtered_df, num_bootstraps = 100)

result_1 = data.frame(
  Variable=c("ATE","SE","CI"),
  Values = c(result_1$ATE, result_1$SE, result_1$CI)
)

kable(result_1, caption = "TRAB: the average treatment effect (ATE)", align = "c")
```

Average Treatment Effect of the Treated due to CMV on TRAB clones
```{r}
# using the function to calculate ATT and CI from the homework 2 key
calculate_ATT_lm <- function(model_formula, data, num_bootstraps = 100) {
 
  model <- lm(model_formula, data = data)
  
  data_trt <- data_con <- data[data$CMV_status == 1, ]
  data_con$CMV_status = 0
  
  pred1 <- predict(model, newdata = data_trt, type = "response")
  pred0 <- predict(model, newdata = data_con, type = "response")
  ATT <- mean(pred1 - pred0, na.rm = TRUE)
  
  # Bootstrap replication for SE
  att_boot <- c()
  set.seed(101)
  for (i in 1:num_bootstraps) {
    data_boot <- data[sample(nrow(data), size = nrow(data), replace = TRUE), ]
    model_boot <- lm(model_formula, data = data_boot)
    
    data_trt_boot <- data_con_boot <- data_boot[data_boot$CMV_status == 1, ]
    data_con_boot$CMV_status = 0
    
    pred1_boot <- predict(model_boot, newdata = data_trt_boot, type = "response")
    pred0_boot <- predict(model_boot, newdata = data_con_boot, type = "response")
    att_boot <- c(att_boot, mean(pred1_boot - pred0_boot, na.rm = TRUE))
  }
  
  # Calculate standard error and confidence interval
  SE <- sd(att_boot)
  CI <- ATT + c(-1, 1) * qnorm(0.975) * SE
  CI_str <- paste0(round(CI[1], 4), ", ", round(CI[2], 4))
  
  return(list(ATT = round(ATT, 4), SE = round(SE, 4), CI = CI_str))
}

# the average treatment effect among the treated (ATT)
model_formula = TRAB ~ CMV_status * age + sex + race + educ + smoke_stat + INFLAM_5 + comorb_idx + Tcells
result_1_1 = calculate_ATT_lm(model_formula, filtered_df, num_bootstraps = 100)

result_1_1 <- data.frame(
  Variable=c("ATT","SE","CI"),
  Values = c(result_1_1$ATT, result_1_1$SE, result_1_1$CI)
)

kable(result_1_1, caption = "TRAB: ATT", align = "c")
```

<br>
<br>


#### **Propensity score model**

```{r}
# removing null rows in covariates
columns_to_check =c('age', 'sex', 'race', 'educ', 'smoke_stat', 'total_readCount', 'BMI', 'INFLAM_5', 'comorb_idx', 'Tcells')
filtered_df2 <- filtered_df %>% 
                  filter(across(all_of(columns_to_check), ~ !is.na(.)))

cat("\nFiltered data size:", nrow(filtered_df2))
```


**Finding the IPW standardized mean difference among treatment groups**

```{r}
# calculating propensity scores and IPW
mod_ps = glm(CMV_status ~ age + sex + race + educ + smoke_stat + INFLAM_5 + comorb_idx + Tcells, family = binomial(), data = filtered_df2)

filtered_df2$ps = predict(mod_ps, type = "response")
filtered_df2$weight = filtered_df2$CMV_status/ filtered_df2$ps + (1 - filtered_df2$CMV_status)/(1 - filtered_df2$ps)

# setting variable names for TableOne
filtered_df2 = labelled::set_variable_labels(filtered_df2,
                                      age = "Age at 2016",
                                      sex = "Biological Sex",
                                      race = "Racial Group",
                                      educ = "Education Level",
                                      smoke_stat = "Smoking Status",
                                      mortality = "Mortality",
                                      TRAB = "T alpha-beta clones",
                                      Tcells = "T cells count",
                                      CD4N_pct = "Naive Helper T cells pct",
                                      BMI = "Body Mass Index",
                                      INFLAM_5 = "Total Inflammation Score",
                                      comorb_idx = "# of comorbidities"
                                      )

# creating weighted survey data and computing SMD
svy_data = svydesign(ids=~1, data=filtered_df2, weights=~weight)

# creating a table one stratified by treatment group
cat_cols = c('sex', 'race', 'smoke_stat', 'mortality')
num_cols = c('age', 'educ', 'BMI', 'INFLAM_5', 'comorb_idx', 'Tcells', 'TRAB', 'CD4N_pct')
all_cols = c(cat_cols, num_cols)
svy_table_one = svyCreateTableOne(vars = all_cols, factorVars = cat_cols, strata = "CMV_status", data=svy_data, test = FALSE)

t2 = print(svy_table_one, nonnormal=num_cols, showAllLevels=FALSE, varLabels = TRUE, smd=TRUE)
```

```{r}
kable(t2, caption = "**IPW weighted participant characteristics stratified by CMV**", align = "c")
```

```{r}
# t2_df <- as.data.frame(t2)
# write.csv(t2_df, "D:/MS/My Research/Plan B/Code/CMV and TCR/ps_weighted_tableone.csv", row.names = TRUE)
```


<br>
<br>


**Estimating the ATE and ATT using propensity score stratification**

```{r}
### For average treatment effect (ATE)
# ps score stratification
ps_quintile = cut(filtered_df2$ps, breaks = quantile(filtered_df2$ps, probs = seq(0, 1, by = 0.2)), include.lowest = TRUE, labels = FALSE)

# finding TE for each quintile
te_quintile = tapply(filtered_df2$TRAB[filtered_df2$CMV_status == 1],
                      ps_quintile[filtered_df2$CMV_status == 1], mean) -  
               tapply(filtered_df2$TRAB[filtered_df2$CMV_status == 0],
                      ps_quintile[filtered_df2$CMV_status == 0], mean)

# finding overall ATE based on ps score stratification
ATE_PSS = sum(te_quintile * table(ps_quintile) / nrow(filtered_df), na.rm = TRUE)

n = nrow(filtered_df)
ATE_PSS_boot = c()
set.seed(101)
# bootstrapping to find standard error
for(i in 1:100) {
  filtered_boot = filtered_df2[sample(1:n, n, replace = TRUE), ]
  mod_ps_boot = glm(CMV_status ~ age + sex + race + educ + smoke_stat + INFLAM_5 + comorb_idx + Tcells,
                 data = filtered_boot, family = "binomial")
  
  ps_boot = predict(mod_ps_boot, type = "response")
  
  # ps score stratification
  ps_quintile_boot = cut(ps_boot, breaks = quantile(ps_boot, probs = seq(0, 1, by = 0.2), na.rm = TRUE), include.lowest = TRUE, labels = FALSE)
  
  # finding TE for each quintile
  te_quintile_boot = tapply(filtered_boot$TRAB[filtered_boot$CMV_status == 1],
                             ps_quintile_boot[filtered_boot$CMV_status == 1], mean) -
                     tapply(filtered_boot$TRAB[filtered_boot$CMV_status == 0],
                             ps_quintile_boot[filtered_boot$CMV_status == 0], mean)
  
  # finding overall ATE based on ps score stratification
  ATE_boot = sum(te_quintile_boot * table(ps_quintile_boot)/ n, na.rm = TRUE)
  
  ATE_PSS_boot <- c(ATE_PSS_boot, ATE_boot)
}

SE_ATE = sd(ATE_PSS_boot)
CI_ATE_PSS = ATE_PSS + c(-1, 1) * qnorm(0.975) * SE_ATE

result_2 = data.frame(
  Variable=c("ATE","SE","lower_CI", "upper_CI"),
  Values = c(ATE_PSS, SE_ATE, CI_ATE_PSS[1], CI_ATE_PSS[2])
)

kable(result_2, caption = "Estimate of ATE using PSS")
```

```{r}
### For average treatment effect among the treated (ATT)
# ps score stratification
ps_quintile = cut(filtered_df2$ps, breaks = c(0, quantile(filtered_df2$ps[filtered_df2$CMV_status == 1], p = c(0.2, 0.4, 0.6, 0.8)), 1), labels = 1:5)

# finding TE of the treated for each quintile
te_quintile = tapply(filtered_df2$TRAB[filtered_df2$CMV_status == 1],
                      ps_quintile[filtered_df2$CMV_status == 1], mean) -  
               tapply(filtered_df2$TRAB[filtered_df2$CMV_status == 0],
                      ps_quintile[filtered_df2$CMV_status == 0], mean)

# finding overall ATT based on ps score stratification
nA = nrow(filtered_df2[filtered_df2$CMV_status == 1, ])
nAj = table(ps_quintile[filtered_df2$CMV_status == 1])

ATT_PSS = sum(te_quintile * nAj / nA, na.rm = TRUE)

n = nrow(filtered_df2)
ATT_PSS_boot = c()
set.seed(101)
# bootstrapping to find standard error
for(i in 1:100) {
  filtered_boot = filtered_df2[sample(1:n, n, replace = TRUE), ]
  mod_ps_boot = glm(CMV_status ~ age + sex + race + educ + smoke_stat + INFLAM_5 + comorb_idx + Tcells,
                 data = filtered_boot, family = "binomial")
  
  ps_boot = predict(mod_ps_boot, type = "response")
  
  # ps score stratification
  ps_quintile_boot = cut(ps_boot, breaks = c(0, quantile(ps_boot[filtered_boot$CMV_status == 1], p = c(0.2, 0.4, 0.6, 0.8)), 1), labels = 1:5)

  # finding TE of the treated for each quintile
  te_quintile = tapply(filtered_boot$TRAB[filtered_boot$CMV_status == 1],
                        ps_quintile_boot[filtered_boot$CMV_status == 1], mean) -  
                 tapply(filtered_boot$TRAB[filtered_boot$CMV_status == 0],
                        ps_quintile_boot[filtered_boot$CMV_status == 0], mean)
  
  # finding overall ATT based on ps score stratification
  nA_boot = nrow(filtered_boot[filtered_boot$CMV_status == 1, ])
  nAj_boot = table(ps_quintile_boot[filtered_boot$CMV_status == 1])
  
  ATT_boot = sum(te_quintile * nAj_boot / nA_boot, na.rm = TRUE)
  
  ATT_PSS_boot = c(ATE_PSS_boot, ATT_boot)
}

SE_ATT = sd(ATT_PSS_boot)
CI_ATT_PSS = ATT_PSS + c(-1, 1) * qnorm(0.975) * SE_ATT

result_2_1 = data.frame(
  Variable=c("ATT","SE","lower_CI", "upper_CI"),
  Values = c(ATT_PSS, SE_ATT, CI_ATT_PSS[1], CI_ATT_PSS[2])
)

kable(result_2_1, caption = "Estimate of ATT using PSS")
```

<br>
<br>

#### **Estimating the ATE and ATT using IPW**

```{r}
# Set seed for reproducibility
set.seed(101)

# trimming extreme ps values
ps = pmin(pmax(filtered_df2$ps, 0.01), 0.99)

# calculate stabilized weights
p_treated = mean(filtered_df2$CMV_status)
w = ifelse(filtered_df2$CMV_status == 1,
            p_treated / ps,
            (1 - p_treated) / (1 - ps))

# compute IPW ATE
Y1 = filtered_df2$TRAB[filtered_df2$CMV_status == 1]
Y0 = filtered_df2$TRAB[filtered_df2$CMV_status == 0]
w1 = w[filtered_df2$CMV_status == 1]
w0 = w[filtered_df2$CMV_status == 0]

ATE_ipw = weighted.mean(Y1, w1, na.rm = TRUE) - 
           weighted.mean(Y0, w0, na.rm = TRUE)

# bootstrapping to find standard error
n = nrow(filtered_df2)
n_boot = 100
ATE_ipw_boot = numeric(n_boot)

for (i in 1:n_boot) {
  idx = sample(1:n, n, replace = TRUE)
  boot_df = filtered_df2[idx, ]
  
  mod_boot = glm(CMV_status ~ age + sex + race + educ + smoke_stat + INFLAM_5 + comorb_idx + Tcells,
                  data = boot_df, family = "binomial")
  ps_boot = predict(mod_boot, type = "response")
  ps_boot = pmin(pmax(ps_boot, 0.01), 0.99)
  
  # calculate stabilized weights
  p_treat_boot = mean(boot_df$CMV_status)
  w_boot = ifelse(boot_df$CMV_status == 1,
                   p_treat_boot / ps_boot,
                   (1 - p_treat_boot) / (1 - ps_boot))
  
  Y1b = boot_df$TRAB[boot_df$CMV_status == 1]
  Y0b = boot_df$TRAB[boot_df$CMV_status == 0]
  w1b = w_boot[boot_df$CMV_status == 1]
  w0b = w_boot[boot_df$CMV_status == 0]
  
  ATE_boot = weighted.mean(Y1b, w1b, na.rm = TRUE) - 
              weighted.mean(Y0b, w0b, na.rm = TRUE)
  ATE_ipw_boot[i] = ATE_boot
}

SE_ATE_ipw = sd(ATE_ipw_boot)
CI_ATE_ipw = ATE_ipw + c(-1, 1) * qnorm(0.975) * SE_ATE_ipw

result3 = data.frame(
  Estimate = c("ATE", "SE", "Lower CI", "Upper CI"),
  Value = round(c(ATE_ipw, SE_ATE_ipw, CI_ATE_ipw), 4)
)

kable(result3, caption = "ATE Estimate using IPW")
```


```{r}
# calculating ATT weights
w_att = ifelse(filtered_df2$CMV_status == 1, 1, ps / (1 - ps))

# computing IPW ATT
Y1 = filtered_df2$TRAB[filtered_df2$CMV_status == 1]
Y0 = filtered_df2$TRAB[filtered_df2$CMV_status == 0]
w0 = w_att[filtered_df2$CMV_status == 0]

ATT_ipw = mean(Y1, na.rm = TRUE) - weighted.mean(Y0, w0, na.rm = TRUE)

# bootstrapping to find standard error
n = nrow(filtered_df2)
n_boot = 100
ATT_ipw_boot = numeric(n_boot)

for (i in 1:n_boot) {
  idx = sample(1:n, n, replace = TRUE)
  boot_df = filtered_df2[idx, ]
  
  mod_boot = glm(CMV_status ~ age + sex + race + educ + smoke_stat + INFLAM_5 + comorb_idx + Tcells,
                  data = boot_df, family = "binomial")
  ps_boot = predict(mod_boot, type = "response")
  ps_boot = pmin(pmax(ps_boot, 0.01), 0.99)
  
  w_att_boot = ifelse(boot_df$CMV_status == 1,
                       1,
                       ps_boot / (1 - ps_boot))
  
  Y1b = boot_df$TRAB[boot_df$CMV_status == 1]
  Y0b = boot_df$TRAB[boot_df$CMV_status == 0]
  w0b = w_att_boot[boot_df$CMV_status == 0]
  
  ATT_boot = mean(Y1b, na.rm = TRUE) - weighted.mean(Y0b, w0b, na.rm = TRUE)
  ATT_ipw_boot[i] = ATT_boot
}

SE_ATT_ipw = sd(ATT_ipw_boot)
CI_ATT_ipw = ATT_ipw + c(-1, 1) * qnorm(0.975) * SE_ATT_ipw

result3_1 = data.frame(
  Estimate = c("ATT", "SE", "Lower CI", "Upper CI"),
  Value = round(c(ATT_ipw, SE_ATT_ipw, CI_ATT_ipw), 4)
)

kable(result3_1, caption = "ATT Estimate using IPW")
```

<br>
<br>


#### **Performing 1:1 nearest neighbor propensity score matching**

**Finding the standardized mean difference among the matched cohort.**

```{r}
# 1 to 1 matching
mod_match_1_1 = matchit(1-CMV_status ~ age + sex + race + educ + smoke_stat + INFLAM_5 + comorb_idx + Tcells, 
                        distance="logit", 
                        method="nearest", 
                        data=filtered_df2, 
                        ratio=1, 
                        replace=TRUE)

match_pairs = na.omit(mod_match_1_1$match.matrix)  # to check if controls re used or not
num_match = nrow(match_pairs)
cat("No of matched rows:", num_match) 

# creating the matched dataframe
matched_df = data.frame(rbind(filtered_df2[filtered_df2$CMV_status == 0, ],
	            filtered_df2[as.numeric(mod_match_1_1$match.matrix), ]))

# setting variable names for TableOne
matched_df = labelled::set_variable_labels(matched_df,
                                      age = "Age at 2016",
                                      sex = "Biological Sex",
                                      race = "Racial Group",
                                      educ = "Education Level",
                                      smoke_stat = "Smoking Status",
                                      mortality = "Mortality",
                                      TRAB = "T alpha-beta clones",
                                      Tcells = "T cells count",
                                      CD4N_pct = "Naive Helper T cells pct",
                                      INFLAM_5 = "Total Inflammation Score",
                                      comorb_idx = "# of comorbidities"
                                      )

# creating a table one stratified by treatment group
cat_cols = c('sex', 'race', 'mortality', 'smoke_stat')
num_cols = c('age', 'educ', 'BMI', 'INFLAM_5', 'comorb_idx', 'Tcells', 'TRAB', 'CD4N_pct')
all_cols = c(cat_cols, num_cols)

table3 = CreateTableOne(data = matched_df, vars=all_cols, factorVars = cat_cols, strata = "CMV_status", test=FALSE)
table3 = print(table3, nonnormal=num_cols, varLabels = TRUE, smd=TRUE)
```

```{r, echo=FALSE, eval=TRUE, include=TRUE, warning=FALSE, message = FALSE, results='hold', cache=TRUE}
kable(table3, caption = "**Participant Characteristics after Matching**", align = "c")
```

<br>

**Propensity score model with the matched data**

```{r}
# calculating propensity scores and IPW
mod_match_ps = glm(CMV_status ~ age + sex + race + educ + smoke_stat + INFLAM_5 + comorb_idx + Tcells, family = binomial(), data = matched_df)

matched_df$ps = predict(mod_match_ps, type = "response")
matched_df$weight = matched_df$CMV_status/ matched_df$ps + (1 - matched_df$CMV_status)/(1 - matched_df$ps)

# creating weighted survey data and computing SMD
svy_data = svydesign(ids=~1, data=matched_df, weights=~weight)

# creating a table one stratified by treatment group
cat_cols = c('sex', 'race', 'mortality', 'smoke_stat')
num_cols = c('age', 'educ', 'BMI', 'INFLAM_5', 'comorb_idx', 'Tcells', 'TRAB', 'CD4N_pct')
all_cols = c(cat_cols, num_cols)
svy_table_one = svyCreateTableOne(vars = all_cols, factorVars = cat_cols, strata = "CMV_status", data=svy_data, test = FALSE)

table4 = print(svy_table_one, nonnormal=num_cols, showAllLevels=FALSE, varLabels = TRUE, smd=TRUE)
```

```{r, eval=TRUE, echo=TRUE, warning=FALSE, message=FALSE, comment=NA, results='hold'}
kable(table4, caption = "**Participant characteristics stratified by IPW**", align = "c")

```


**Estimating the ATE and ATT using propensity score stratification**

```{r}
### For average treatment effect (ATE)
# ps score stratification
ps_quintile = cut(matched_df$ps, breaks = quantile(matched_df$ps, probs = seq(0, 1, by = 0.2)), include.lowest = TRUE, labels = FALSE)

# finding TE for each quintile
te_quintile = tapply(matched_df$TRAB[matched_df$CMV_status == 1],
                      ps_quintile[matched_df$CMV_status == 1], mean) -  
               tapply(matched_df$TRAB[matched_df$CMV_status == 0],
                      ps_quintile[matched_df$CMV_status == 0], mean)

# finding overall ATE based on ps score stratification
ATE_PSS = sum(te_quintile * table(ps_quintile) / nrow(matched_df), na.rm = TRUE)

n = nrow(matched_df)
ATE_PSS_boot = c()
set.seed(101)
# bootstrapping to find standard error
for(i in 1:100) {
  matched_boot = matched_df[sample(1:n, n, replace = TRUE), ]
  mod_ps_boot = glm(CMV_status ~ age + sex + race + educ + smoke_stat + INFLAM_5 + comorb_idx + Tcells,
                 data = matched_boot, family = "binomial")
  
  ps_boot = predict(mod_ps_boot, type = "response")
  
  # ps score stratification
  ps_quintile_boot = cut(ps_boot, breaks = quantile(ps_boot, probs = seq(0, 1, by = 0.2), na.rm = TRUE), include.lowest = TRUE, labels = FALSE)
  
  # finding TE for each quintile
  te_quintile_boot = tapply(matched_boot$TRAB[matched_boot$CMV_status == 1],
                             ps_quintile_boot[matched_boot$CMV_status == 1], mean) -
                     tapply(matched_boot$TRAB[matched_boot$CMV_status == 0],
                             ps_quintile_boot[matched_boot$CMV_status == 0], mean)
  
  # finding overall ATE based on ps score stratification
  ATE_boot = sum(te_quintile_boot * table(ps_quintile_boot)/ n, na.rm = TRUE)
  
  ATE_PSS_boot <- c(ATE_PSS_boot, ATE_boot)
}

SE_ATE = sd(ATE_PSS_boot)
CI_ATE_PSS = ATE_PSS + c(-1, 1) * qnorm(0.975) * SE_ATE

result_4 = data.frame(
  Variable=c("ATE","SE","lower_CI", "upper_CI"),
  Values = c(ATE_PSS, SE_ATE, CI_ATE_PSS[1], CI_ATE_PSS[2])
)

kable(result_4, caption = "Estimate of ATE using PSS")
```


```{r}
### For average treatment effect among the treated (ATT)
# ps score stratification
ps_quintile = cut(matched_df$ps, breaks = c(0, quantile(matched_df$ps[matched_df$CMV_status == 1], p = c(0.2, 0.4, 0.6, 0.8)), 1), labels = 1:5)

# finding TE of the treated for each quintile
te_quintile = tapply(matched_df$TRAB[matched_df$CMV_status == 1],
                      ps_quintile[matched_df$CMV_status == 1], mean) -  
               tapply(matched_df$TRAB[matched_df$CMV_status == 0],
                      ps_quintile[matched_df$CMV_status == 0], mean)

# finding overall ATT based on ps score stratification
nA = nrow(filtered_df[matched_df$CMV_status == 1, ])
nAj = table(ps_quintile[matched_df$CMV_status == 1])

ATT_PSS = sum(te_quintile * nAj / nA, na.rm = TRUE)

n = nrow(matched_df)
ATT_PSS_boot = c()
set.seed(101)
# bootstrapping to find standard error
for(i in 1:100) {
  matched_boot = matched_df[sample(1:n, n, replace = TRUE), ]
  mod_ps_boot = glm(CMV_status ~ age + sex + race + educ + smoke_stat + INFLAM_5 + comorb_idx + Tcells,
                 data = matched_boot, family = "binomial")
  
  ps_boot = predict(mod_ps_boot, type = "response")
  
  # ps score stratification
  ps_quintile_boot = cut(ps_boot, breaks = c(0, quantile(ps_boot[matched_boot$CMV_status == 1], p = c(0.2, 0.4, 0.6, 0.8)), 1), labels = 1:5)

  # finding TE of the treated for each quintile
  te_quintile = tapply(matched_boot$TRAB[matched_boot$CMV_status == 1],
                        ps_quintile_boot[matched_boot$CMV_status == 1], mean) -  
                 tapply(matched_boot$TRAB[matched_boot$CMV_status == 0],
                        ps_quintile_boot[matched_boot$CMV_status == 0], mean)
  
  # finding overall ATT based on ps score stratification
  nA_boot = nrow(matched_boot[matched_boot$CMV_status == 1, ])
  nAj_boot = table(ps_quintile_boot[matched_boot$CMV_status == 1])
  
  ATT_boot = sum(te_quintile * nAj_boot / nA_boot, na.rm = TRUE)
  
  ATT_PSS_boot = c(ATE_PSS_boot, ATT_boot)
}

SE_ATT = sd(ATT_PSS_boot)
CI_ATT_PSS = ATT_PSS + c(-1, 1) * qnorm(0.975) * SE_ATT

result_4_1 = data.frame(
  Variable=c("ATT","SE","lower_CI", "upper_CI"),
  Values = c(ATT_PSS, SE_ATT, CI_ATT_PSS[1], CI_ATT_PSS[2])
)

kable(result_4_1, caption = "Estimate of ATT using PSS")
```

<br>
<br>




